---
title: "Bioinformatics class7"
author: "Katie Wall"
date: "24 April 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions again
Here I am going to revisit our function from class 6

```{r}
source("http://tinyurl.com/rescale-R")
```
lets see if we can use this function
```{r}
rescale(1:10)
```

looks good lets break it

```{r}
rescale( c(1:10,"string"))
```
lets try the new **rescale2()** function
```{r}
rescale2( c(1:10,"string"))
```

## Write an NA checking function
Here we write a new function to check for NAs in two inputs
```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)

```

```{r}
which( is.na(x))
```

```{r}
sum( is.na(x))
```


lets put this together with x and y
```{r}
is.na(x)
is.na(y)
```

```{r}
is.na(x) & is.na(y)
```


```{r}
sum( is.na(x) & is.na(y) )
```

This is the both_na() function startting point
```{r}
both_na <- function(x, y) {
  ## Check for NA elements in both input vectors 
  sum( is.na(x) & is.na(y) )
}
```
```{r}
both_na(x,y)
```

and again 
```{r}
x <- c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)

both_na(x, y2)
```

use stop() inside function to catch this case 
```{r}
both_na2(x, y2)
```


Try the both_na3() function
```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)

ans <- both_na3(x, y)
```

```{r}
ans$number
```


## Another function example: gene intersection

```{r}
df1
df2

x <- df1$IDs
y <- df2$IDs
```
```{r}

x
y
```

```{r}
intersect(x, y)
```

```{r}
#intersect(x, y)

x %in% y
```

```{r}
y %in% x
```

Now we can access the genes we want with these 'indices'

```{r}
x[x %in% y]
y[y %in% x]
```

can makwe these columns of the same object with **cbind()**
```{r}
cbind(x[x %in% y], y[y %in% x])
```



```{r}
# Correct but yucky format for 2nd colnames
gene_intersect2(df1, df2)
```

lets try merge() function for this job

```{r}
merge(df1, df2, by="IDs")
```



